Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm':' A Case Study

### 2022-04-13 Paper_Draft2/Draft2.tex
\\textbf{}

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
debiasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library. Our project
builds on Women at the Table's various debiasing algorithms used in its
AI \& Equality Human Rights Toolbox to conduct our own analyses on the
COMPAS dataset. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various debiasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores dataset from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial preprocessing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race, age (Figure
\ref{fig:age plot}), and two year recidivism rate (Figures
\ref{fig:recid plot} \& \ref{fig:recid race plot}).

@@

\emph{Statistical Parity Difference} - This metric measures the
difference that the majority and protected classes get a particular
outcome. The ideal value of this metric is 0. Fairness for this metric
is between -0.1 and 0.1. A negative value means there is higher benefit
for the privileged group (in this case, white defendants).

@@

\emph{Disparate Impact Ratio} - This metric is the ratio of how often
the favorable outcome occurs in one group versus the other. In the case
of recidivism, this is the ratio of how many white defendants are
predicted to not recidivate compared to how many black defendants are
predicted to not recidivate. A value of 1 means that the ratio is
exactly 1:1. Less than 1 means the privileged group (white defendants)
benefits, while a value greater than 1 means the unprivileged group
(Black defendants) benefits. According to AIF360, a ratio between .8 to
1.25 is considered fair:

@@

\emph{Average Odds Difference} - This metric returns the average
difference in false positive rate and true positive rate for the
privileged and unprivileged groups. A value of 0 indicates equality of
odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

@@

\emph{Equal Opportunity Difference} - This metric is computed as the
difference of true positive rates between the unprivileged and the
privileged groups. The true positive rate is the ratio of true positives
to the total number of actual positives for a given group.

The ideal value is 0. A value less than 0 implies higher benefit for the
privileged group and a value greater than 0 implies higher benefit for
the unprivileged group. Fairness for this metric is between -0.1 and
0.1. This metric is best used when it is very important to catch
positive outcomes while false positives are not exceptionally
problematic. This is not the case for the COMPAS data set, as false
positives mean extra jail time for someone who will not actually
reoffend.

@@

For the next step of our experiment, we needed to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing debiasing algorithms.
These are all based on using predictive models to figure out how we can
``fix'' the bias that is present. Pre-processing refers to assessing the
training data, and it is the most flexible method because it has not yet
trained a model that may carry assumptions about the data. It is
important to keep in mind that pre-processing prevents assumptions in
the modeling, but does not account for the bias in data collection. With
pre-processing, we use the reweighting algorithm from AIF360 which
weights the examples in each (group, label) combination differently to
ensure fairness before classification \citep{aif360-oct-2018}. We used a
logistic regression model for this algorithm, as it is the easiest to
interpret in the given context. After running the fairness metrics using
the pre-processing algorithm, we were able to compare our results to the
baseline metrics from the previous section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. In order to do this, we first needed to convert our data frame
into a data type called a BinaryLabelDataset. Similarly to
pre-processing, we compared the results of our in-processing methods
with both the baseline and the pre-processing to gauge which method so
far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and uninterpretable, post-processors are trained in a
black box. To get around this, we use the PostProcessingMeta class from
AIF360 which combines the training and prediction for a random estimator
and the post-processor while simultaneously splitting the dataset. Based
on these results, we looked comparatively at all the approaches to
determine which we assessed to be the most fair.

>>>>

\href{https://www.womenatthetable.net/}{Women at the Table}, the sponsor
organization for this project, is ``a growing, global gender equality \&
democracy CSO based in Geneva, Switzerland focused on advancing feminist
systems change by using the prism of technology, innovation \& AI
exercising leverage points in technology, the economy, sustainability \&
democratic governance.'' We are collaborating with the organization on
its AI \& Equality \citep{noauthor_ai_nodate} initiative, tasked with
de-biasing the COMPAS algorithm \citep{aif360-oct-2018} and producing a
corresponding data story that will be added to its library. Our project
builds on Women at the Table's various de-biasing algorithms used in its
AI \& Equality Human Rights Toolbox to conduct our own analyses on the
COMPAS data set. Based on this analysis, we employ a human rights
framework to contribute to the ProPublica and Northpointe debate and
investigate whether or to what extent there is racial bias in the COMPAS
algorithm. With a solid understanding of the two sides, we aim to
pinpoint the shortcomings of both arguments and correct them in our
analyses. We will use various de-biasing techniques and fairness metrics
to evaluate the level of bias present in the COMPAS data and our
algorithm. We will summarize our results using the JupyterNotebook
framework from Women at the Table, to be used by members of the
organization to teach in a workshop setting. We hope that our findings
will highlight the importance of checking statistical analyses using
varied methods and contribute to the ongoing discussion of the effects
of machine biases in the justice system.

@@

The data we are using for this project is the COMPAS General Recidivism
Risk Scores data set from the AI Fairness 360 (AIF360) toolkit
\citep{aif360-oct-2018}, which does the same initial pre-processing as
ProPublica. The raw data has 6,167 rows and each row represents an
arrest charge for a defendant. AIF360's COMPAS data includes the
defendant's age, race, sex, what they were charged with, and whether or
not the defendant ultimately recidivated within a two-year period after
their arrest. For the purposes of our project, which endeavors to
evaluate the differing effects of the COMPAS algorithm on white
defendants versus Black defendants, we have filtered the data to only
include individuals whose race is listed as Caucasian or
African-American. Our data therefore has 5,723 rows (Figure
\ref{fig:table snip}), with the below distributions of race, age (Figure
\ref{fig:age plot}), and two year recidivism rate (Figures
\ref{fig:recid plot} \& \ref{fig:recid race plot}).

@@

\textbackslash textbf\{Statistical Parity Difference\} - This metric
measures the difference that the majority and protected classes get a
particular outcome. The ideal value of this metric is 0. Fairness for
this metric is between -0.1 and 0.1. A negative value means there is
higher benefit for the privileged group (in this case, white
defendants).

@@

\textbackslash textbf\{Disparate Impact Ratio\} - This metric is the
ratio of how often the favorable outcome occurs in one group versus the
other. In the case of recidivism, this is the ratio of how many white
defendants are predicted to not recidivate compared to how many black
defendants are predicted to not recidivate. A value of 1 means that the
ratio is exactly 1:1. Less than 1 means the privileged group (white
defendants) benefits, while a value greater than 1 means the
unprivileged group (Black defendants) benefits. According to AIF360, a
ratio between .8 to 1.25 is considered fair:

@@

\textbackslash textbf\{Average Odds Difference\} - This metric returns
the average difference in false positive rate and true positive rate for
the privileged and unprivileged groups. A value of 0 indicates equality
of odds, and a value below 0 implies benefit for the privileged group.
Equality of odds is achieved in the case of recidivism when the
proportion of people who were predicted to recidivate and did recidivate
is equal (true positive rate) for both Black and white defendants AND
the proportion of people who were predicted to recidivate and did not
recidivate (false positive rate) is equal for both Black and white
defendants.

@@

\textbackslash textbf\{Equal Opportunity Difference\} - This metric is
computed as the difference of true positive rates between the
unprivileged and the privileged groups. The true positive rate is the
ratio of true positives to the total number of actual positives for a
given group.

The ideal value is 0. A value less than 0 implies higher benefit for the
privileged group and a value greater than 0 implies higher benefit for
the unprivileged group. Fairness for this metric is between -0.1 and
0.1. This metric is best used when it is very important to catch
positive outcomes while false positives are not exceptionally
problematic. This is not the case for the COMPAS data set, as false
positives mean extra jail time for someone who will not actually
re-offend.

@@

For the next step of our experiment, we needed to determine where in the
data science pipeline we can mitigate the most bias, using
pre-processing, in-processing, and post-processing de-biasing
algorithms. These are all based on using predictive models to figure out
how we can ``fix'' the bias that is present. Pre-processing refers to
assessing the training data, and it is the most flexible method because
it has not yet trained a model that may carry assumptions about the
data. It is important to keep in mind that pre-processing prevents
assumptions in the modeling, but does not account for the bias in data
collection. With pre-processing, we use the re-weighing algorithm from
AIF360 which weights the examples in each (group, label) combination
differently to ensure fairness before classification
\citep{aif360-oct-2018}. We used a logistic regression model for this
algorithm, as it is the easiest to interpret in the given context. After
running the fairness metrics using the pre-processing algorithm, we were
able to compare our results to the baseline metrics from the previous
section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. In order to do this, we first needed to convert our data frame
into a data type called a . Similarly to pre-processing, we compared the
results of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and difficult to interpret, post-processors are trained
in a black box. To get around this, we use the PostProcessingMeta class
from AIF360 which combines the training and prediction for a random
estimator and the post-processor while simultaneously splitting the data
set. Based on these results, we looked comparatively at all the
approaches to determine which we assessed to be the most fair.
